---
title: "Working with SADCAT descriptors in AGRUPA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with SADCAT descriptors in AGRUPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Overview

This vignette documents the helper functions defined in **Data.R** and **Dictionary.R** used in the AGRUPA project to:

1. Extract candidate descriptors from Spanish text (either *comma-separated* descriptors or *narrative* text).
2. Normalize them to match SADCAT's lookup format (lowercase, whitespace removal, optional diacritic removal).
3. Compute global and dimension-level **coverage** against SADCAT, and compute **mean direction** scores for each dimension/facet.

The pipeline is designed around the **Stereotype Content Model (SCM)** dimensions (warmth and competence) plus additional facets provided by SADCAT.

> **Important:** normalization in this project removes *all whitespace* (e.g., `"alta calidad"` → `"altacalidad"`). This matches the format expected by the SADCAT dictionaries used here. Always verify this assumption by inspecting `SADCAT::Spanishdicts$Palabra`.

## Dependencies

The functions rely on:

- `SADCAT` (dictionary)
- `stopwords` (optional stopword removal)
- `udpipe` (optional lemmatization)
- `stringi` (optional diacritic stripping)

```{r deps}
# install.packages("remotes")
# remotes::install_github("gandalfnicolas/SADCAT", dependencies = TRUE)
# install.packages(c("udpipe", "stopwords", "stringi", "dplyr"))

library(SADCAT)
# Load the package being documented
library(AGRUPA)
```

## Loading the dictionary

```{r dict}
# Spanish dictionary shipped with SADCAT
spanish_dict <- SADCAT::Spanishdicts
names(spanish_dict)[1:10]
```

## 1) Preparing descriptors

### 1.1 Comma-separated descriptors

Use `prepare_descriptors(..., input_format = "comma")` when your input is a string such as:
`"elegante, solemne, aristocrático"`.

```{r comma}
# Source descriptors as a vector of comma-separated strings (one per row)
v <- c(
  "elegante, solemne, aristocrático",
  "barroco, realista"
)

df_desc <- prepare_descriptors(
  x = v,
  input_type = "vector",
  input_format = "comma",
  remove_diacritics = TRUE,
  keep_enye = FALSE
)

df_desc
```

#### Optional: lemmatization for comma descriptors

If you want to add lemma-based candidates, set `lemmatize = "lemma"` or `lemmatize = "both"` and pass an UDPipe model.

```{r comma-lemma, eval=FALSE}
library(udpipe)

m <- udpipe_download_model(language = "spanish-ancora", overwrite = FALSE)
ud_model <- udpipe_load_model(m$file_model)

df_desc_lem <- prepare_descriptors(
  x = v,
  input_type = "vector",
  input_format = "comma",
  lemmatize = "both",
  udpipe_model = ud_model
)
```

### 1.2 Narrative text (unigrams + optional bi/tri-grams)

Use `prepare_descriptors(..., input_format = "text")` when your input is running text
(e.g., a museum catalogue description).

By default, the function extracts:
- **unigrams** (optionally removing stopwords), and
- **bigrams + trigrams** (without stopword removal; segmentation prevents crossing punctuation).

```{r text}
text_vec <- c(
  "Las Meninas es una obra compleja y magistral. Representa a la infanta y su séquito.",
  "Un retrato sobrio y ceremonial, con gran riqueza compositiva."
)

df_text_desc <- prepare_descriptors(
  x = text_vec,
  input_type = "vector",
  input_format = "text",
  include_ngrams = TRUE,
  remove_stopwords = TRUE,
  stopwords_lang = "es",
  remove_diacritics = TRUE,
  keep_enye = FALSE
)

df_text_desc[1, 1:12]
```

#### Optional: lemmatization for narrative text

For narrative text, lemmatization is done **with context** via `udpipe_annotate` on the whole document.
This typically improves lemma quality compared to lemmatizing tokens in isolation.

```{r text-lemma, eval=FALSE}
library(udpipe)

m <- udpipe_download_model(language = "spanish-ancora")
ud_model <- udpipe_load_model(m$file_model)

df_text_desc_lem <- prepare_descriptors(
  x = text_vec,
  input_type = "vector",
  input_format = "text",
  include_ngrams = TRUE,
  remove_stopwords = TRUE,
  stopwords_lang = "es",
  lemmatize = "both",
  udpipe_model = ud_model
)
```

## 2) Coverage against SADCAT

Once you have a data frame with columns starting with `descriptor_`, you can compute coverage.

### 2.1 Global coverage

```{r cov-global}
cov_global <- dict_coverage(df_text_desc, dict = SADCAT::Spanishdicts$Palabra, prefix = "descriptor_")
cov_global[, c("cov_pct_global", "n_descriptores_fila", "n_en_diccionario_fila")]
```

### 2.2 Coverage by dimension/facet

SADCAT encodes membership in dimensions/facets using columns ending in `_dict`. The helper
`dict_dim_coverage_all()` computes the percentage of your candidate descriptors that fall into each of those columns.

```{r cov-dim}
cov_dim <- dict_dim_coverage_all(df_text_desc, dict_df = SADCAT::Spanishdicts, prefix = "descriptor_")

# Show only the new coverage columns:
cov_cols <- grep("^cov_.*_pct$", names(cov_dim), value = TRUE)
cov_dim[1, cov_cols[1:min(8, length(cov_cols))]]
```

## 3) Mean direction scores

SADCAT also provides direction scores in columns ending in `_dir`. The helper
`dict_dim_dirmean_all()` computes the **mean direction** per row (based only on in-vocabulary descriptors with non-missing direction scores),
and also returns the number of contributing descriptors for each mean.

```{r dirmean}
dirmean <- dict_dim_dirmean_all(df_text_desc, dict_df = SADCAT::Spanishdicts, prefix = "descriptor_")

dir_cols <- grep("^dirmean_", names(dirmean), value = TRUE)
n_cols   <- grep("^n_dirmean_", names(dirmean), value = TRUE)

dirmean[1, c(dir_cols[1:min(6, length(dir_cols))], n_cols[1:min(6, length(n_cols))])]
```

## 4) Filtering by coverage

A simple helper is provided to filter rows using a coverage column and a threshold expressed as a percentage.

```{r filter}
# Keep only rows with at least 30% global coverage:
subset_30 <- filter_by_coverage(cov_global, "cov_pct_global", thr_pct = 30)
subset_30
```

## Notes on interpretation

- **Coverage** is a property of *your candidate descriptor set* and SADCAT's lexical coverage.
  A low coverage score does *not* necessarily imply weak stereotype content in the text; it can also reflect
  lexical mismatch (e.g., named entities, rare words, or domain-specific vocabulary).
- **Direction means** are computed only from descriptors that have non-missing direction values in the dictionary.
  Always inspect the accompanying `n_dirmean_*` columns.



